Building and In-Place Checking Suffix Array in External Memory

### 2017-04-06 draft-new/draft.tex
2017/04/07

Suffix array~(SA) can be built within linear time and space by means of the induced sorting~(IS) method. The recently proposed IS suffix sorter fSAIS has a performance competitive to that of the state-of-the-art parallel construction algorithms in external memory, which indicates a great potential for improving the other disk-based IS alternatives by engineering them carefully. In view of this, we redesign the reduction phase of DSA-IS by using a new substring sorting and naming method and implement the variant, called DSA-IS+, by using several space-saving techinques previously applied to fSAIS.

A constructed SA should be checked to ensure its correctness. We present in this paper a method that enables any IS suffix sorter to perform building and checking simultaneously, where the consumption for checking is negligible in comparison with that for building in our experiments.

@@

The suffix array~(SA)~\cite{Manber1993} is an essential data structure for string processing and information retrieval. Among the existing internal-memory algorithms for SA construction, SA-IS~\cite{Nong11} achieves the optimal time and space complexities using the induced sorting principle, where the key operation is to retrieve the preceding characters of sorted suffixes in order for inducing the lexical order of unsorted ones. To handle massive datasets , several external memory algorithms have been proposed for building massive suffix arrays in recent years, e.g., DC3~\cite{Dementiev2008a}, bwt-disk~\cite{Ferragina2012}, SAscan~\cite{Karkkainen2014}, pSAscan~\cite{Karkkainen2015}, eSAIS~\cite{Bingmann2013}, EM-SA-DS~\cite{Nong2014} and DSA-IS~\cite{Nong2015}. Among them, the latter three algorithms are based on the induced sorting~(IS) method described in SA-IS~\cite{Nong2011}.

 This algorithm employs the IS method to induce the lexical order of all the suffixes from a selected subset.

 This algorithm uses the induced sorting method to

In the past decades, great effort has been taken to study efficient algorithms for SA construction. th

extensive works have been put on designing time and space efficient suffix sorting algorithms

data structure that has been widely used in many string processing applications, e.g., biological sequence alignment, time series analysis and text clustering. Given an input string, traversing its suffix tree can be emulated by using the corresponding enhanced suffix array~\cite{Abouelhodaa2004}, which mainly consists of the suffix and the longest common prefix arrays. It has been realized that the application scope of an index mainly depends on the construction speed and the space requirement. This leads to intensive works on designing time and space efficient suffix sorting algorithms over the past decade, assuming different computation models such as internal memory, external memory, parallel and distributed models. Particularly,

The basic idea behind the induced sorting method is to induce the lexicographical order of all the substrings/suffixes from a sorted subset of substrings/suffixes. Following the idea, an IS-based suffix sorting algorithm is typically comprised of a reduction phase for sorting and naming substrings to reduce a string $x[0,n)$ to a short string $x1[0,n1)$ with $n1 \le \frac{1}{2}n$ and an induction phase for sorting suffixes to induce ${\sf SA}(x)$ from ${\sf SA}(x1)$. During the two phases, the key operation is to retrieve the preceding character of a sorted substring/suffix. This can be done very quickly when $x$ is fully accommodated in the internal memory, but will become slow when $x$ resides in the external memory, as each operation takes a random disk access. For a high I/O efficiency, eSAIS, EM-SA-DS and DSA-IS use different auxiliary data structures to retrieve the preceding characters in a disk-friendly way. Particularly, both eSAIS and EM-SA-DS split a long substring into pieces and represent each piece by a fixed-size tuple, while DSA-IS does not. With an elaborate arrangement of the I/O operations, the programs for these three algorithms are competitive with those for others in terms of both time and space efficiencies.

Among the existing internal-memory suffix sorters, SA-IS~\cite{Nong11} achieves the optimal time and space complexities using the induced sorting method, the key operation of which is to retrieve the preceding characters of sorted suffixes in order for inducing the lexical order of unsorted ones. In the past five years, several works adapted SA-IS to design suffix sorters specific for EM models~\cite{Nong14, Nong15, Karkkainen2014, Liu2015}. These variants use different approaches to retrieve the preceding characters by sequential I/O operations. However, they all suffer from a space bottleneck for taking at least twice disk space as SA-IS on real-world datasets. Recently, it was presented in~\cite{xxx} a new engineering version of SA-IS that achieves nearly optimal space efficiency, indicating a great potential for improving the other disk-based IS alternatives by engineering them carefully.

Spae

\section{Conclusion}
xxx

>>>>

The induced sorting~(IS) method has been used to design disk-based algorithms for suffix array~(SA) construction. A recent engineering of these algorithms can achieve nearly optimal disk space and has a higher scalability over the state-of-the-art suffix sorters in terms of time and I/O complexities. In this paper, we propose a checking method that enables any IS suffix sorting algorithms to verify an SA when it is being built. For performance analysis, we integrate the checking method into the IS suffix sorting algorithm DSA-IS and implement the adapted algorithm to investigate the checking overhead. The experimental results demonstrate that the time, space and I/O consmputions for verification is negligible in comparison with that for construction, indicating that the combination of the proposed checking and the IS building methods can constitute an efficient solution for the situations where building and checking must be done simultaneously.

@@

For any text drawn from a constant or integer alphabet, its suffix array can be built on RAM in linear time and space by the SA-IS algorithm~\cite{Nong11}, which adopts the IS method to induce the lexical order of unsorted suffixes from those already sorted. In the past five years, the IS method has been reused to design three disk-based suffix sorting algorithms~eSAIS~\cite{Bingmann12}, DSA-IS~\cite{Nong15} and SAIS-PQ~\cite{Liu15} that employ different approaches to retrieve the heading characters of unsorted suffixes by efficient I/O operations. These algorithms are competitive with the existing alternatives in terms of time and I/O complexities, but their naive implementations consume around 20 times the input size, leading to a performance bottleneck. A	 recent work in~\cite{xxx} presented a careful engineering of these algorithms, which takes less than $8n$ disk space for an input consisting of $n \le 2^{40}$ characters and runs faster than the state-of-the-art suffix sorter pSAscan~\cite{Karkkainen2015} when $n$ is considerably larger than the available RAM space. This reveals that the IS method can serve as a basis for potentially fastest suffix sorting solutions on EM models.

An SA should be verified before use to detect potential computation errors caused by hardware malfunctions or implementation bugs. Currently, the existing SA checker performs two runs of integer sorts and runs slowly in external memory. We propose in this paper a lightweight checking method that enables any disk-based IS suffix sorting algorithms to verify an SA when it is being built. From our experiments, the time, space and I/O volume for verification is negligible in comparison with that for construction.

The rest of this paper is organized as following. We first introduce the preliminaries for this paper in Section~\ref{sec:preliminaries} and describe the idea of the proposed checking method in Section~\ref{sec:idea}. Then, we demonstrate our experimental results in Section~\ref{sec:experiments} and conclude the remarks in Section~\ref{sec:conclusion}.

\section{Preliminaries}~\label{sec:preliminaries}

\section{Checking Idea}~\label{sec:idea}

\section{Experiments}~\label{sec:experiments}

\section{Conclusion}~\label{sec:conlusion}
